Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-04-28 hazelnut-popl17.tex
updated abstract

\maketitle
\begin{abstract}
Programmers typically construct and manipulate well-typed programs indirectly, as text that must pass through a parser and typechecker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer makes a mistake, is commonly found to be malformed or ill-typed. Contending with (i.e. reasoning about) malformed text or well-formed but meaningless syntax trees is a difficult task, for programmers and  tools alike.

@@

We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the program in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a foundation for \emph{live programming environments}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

>>>>

\maketitle
\begin{abstract}
Well-typed expressions are meaningful inductive structures, but programmers typically construct and manipulate them indirectly, as text that must pass through a parser and typechecker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. Contending with (i.e. reasoning about) malformed text or well-formed but meaningless expressions is a difficult task, for programmers and their tools alike.

@@

We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}
